# The model specification

Models are specified as a C++ `class`/`struct` that should have the following
signature

```{cc, echo=TRUE, eval=FALSE}
struct model{
  // data statements
  void preProcess(){
    // preprocess (called only once) here
  } 

  template < class varType, class tensorType, bool storeNames>
  void operator()(amt::amtModel<varType,tensorType,storeNames> &model__){
    // specifcation of model target density ++ here
  } 
}; // end of struct
```

The name of the `class`/`struct` is arbitrary

## A simple worked example {#simple-worked-example} 

Now consider specifying the model $y_i \sim$ iid $N(\mu,\exp(\lambda)),\;i=1,\dots,n$ (with
flat priors on $\mu,\lambda$) for a given data set $\mathbf y$. We wish to sample from the posterior distribution of $(\mu,\lambda)$, and in addition, get samples from the posterior distribution of $\sigma=\exp(0.5\lambda)$.

The model class would look something like
```{cc,echo=TRUE,eval=FALSE}
using namespace amt;
struct model{
  
  DATA_VECTOR(y); // data to be passed from R
  
  void preProcess(){} // not used in this example 

  template < class varType, class tensorType, bool storeNames>
  void operator()(amt::amtModel<varType,tensorType,storeNames> &model__){
    
    PARAMETER_SCALAR(mu); // parameter (sampled quantity)
    PARAMETER_SCALAR(lambda); // parameter (sampled quantity)
    // note; all variables depending on the parameters must be of type varType
    varType sigma = exp(0.5*lambda);
    // add data likelihood to the model object
    model__+=normal_ld(y,mu,sigma);
    // add sigma as a quantity to produce samples of
    model__.generated(sigma,"sigma");
  } 
}; // end of struct
```

When stored in file `basic_model.cpp`, the above model specification may be compiled using

```{r,echo=TRUE,eval=eval.examples}
model <- pdmphmc::build("basic_model.cpp")
```

Then let's simulate some data and run the model:
```{r,echo=TRUE,eval=eval.examples}
set.seed(123)
y <- rnorm(10) # y is iid N(0,1) with n=10
fit <- pdmphmc::run(model,data=list(y=y))
```

Finally, get a summary of the sampled- and generated quantities, based on `rstan::monitor`:
```{r,echo=TRUE,eval=eval.examples}
fit
```

Further functions exist for inspecting the output, e.g. trace plots
```{r,echo=TRUE,eval=eval.examples}
pdmphmc::trace.plot(fit,"sigma")
```

In the next chapter, a more detailed summary of the possibilities of the model specification is given.